### 2024-07-07 research-methods.qmd
20240707 - bidirectional effect

A third possibility is that the association reflects a bidirectional effect, where `X` causes `Y` and `Y` causes `X`, as depicted in INSERT.
For instance, consumption of more sports drink may improve players' performance, and greater performance in turn may lead players to drink more sports drink.

A fourth possibility is that the association could reflect the influence of a third variable.
If a third variable is a common cause of each and accounts for their association, it is a *confound*.
An observed association between `X` and `Y` could reflect a confound—i.e., a cause (`Z`) that influences both `X` and `Y`, which explains why `X` and `Y` are correlated even though they are not causally related.
A third variable confound that is a common cause of both `X` and `Y` is depicted in @fig-correlationAndCausation3.
For instance, it may not be that sport drink consumption per se influences player performance; rather, it may be that players who are more intelligent or have more financial resources tend to drink more sports drinks and also tend to perform better.
In this case, intelligence or financial resources may be a confound that influences both sports drink consumption and player performance, but sports drink consumptions—though correlated with player performance—does not influence player performance.

@@

>>>>

A third possibility is that the association reflects a bidirectional effect, where `X` causes `Y` and `Y` causes `X`, as depicted in @fig-correlationAndCausation3.
For instance, consumption of more sports drink may improve players' performance, and greater performance in turn may lead players to drink more sports drink.

Bidirectional Effect Between `X` and `Y`, such that `X` causes `Y` and `Y` causes `X`.
:::

A fourth possibility is that the association could reflect the influence of a third variable.
If a third variable is a common cause of each and accounts for their association, it is a *confound*.
An observed association between `X` and `Y` could reflect a confound—i.e., a cause (`Z`) that influences both `X` and `Y`, which explains why `X` and `Y` are correlated even though they are not causally related.
A third variable confound that is a common cause of both `X` and `Y` is depicted in @fig-correlationAndCausation4.
For instance, it may not be that sport drink consumption per se influences player performance; rather, it may be that players who are more intelligent or have more financial resources tend to drink more sports drinks and also tend to perform better.
In this case, intelligence or financial resources may be a confound that influences both sports drink consumption and player performance, but sports drink consumptions—though correlated with player performance—does not influence player performance.

@@

::: {#fig-correlationAndCausation4}
![](images/correlationAndCausation_4.png)
